---
title: "Assignment 6 - Open data and API"
author: "Meghan Stewart"  
format:
  html:
    code-fold: true 
    embed-resources: true
    toc: true
    number-sections: true
    theme: cerulean
---

# Goal of this assignment

The goal of this assignment is for you to leverage open data APIs in R related to your field of study.

In specific, you will be asked to:\
- Identify an open data base related to your field of study\
- Retrieve data from it into R (bonus points if through an API instead of manually)\
- Create questions one could ask of these data, and answer them with wrangling, tables, and plots

# Instructions

1.  Identify an open data in relation to your field of study. The minimum requirements for these data are:\

-   It needs to be open (i.e., public data that anyone can access)\
-   It needs to report some variable of importance across time (e.g., multiple days, months, years, etc.) and/or space (e.g., cities, counties, states, countries, etc.), ideally both.

2.  Bring these data into R, bonus points if you use an API for that (instead of manually)

3.  Create 6 questions you can ask of these data. Of these questions:\

-   At least 4 should require some wrangling to be answered (like a group_by and summarise, etc.)\
-   At least 3 should be answered with a graph

```{r FDA API library Chunk}
#| warning: false
library(jsonlite)
library(httr)
library(ggplot2)
library(tidyverse)
library(dplyr)
library(purrr)

```

# Questions

## 1. What is your field of study?

**Food Science & Technology**

## 2. What is the open data you identified? Please explain it in terms of its source (website, which agency reported it), variable of importance, temporal and spatial characteristics, and anything else relevant for someone outside of your field of study to fully understand what it represents.

***The open data source identified is from the FDA and is a collection of reported adverse events from food products in the U.S. The particular dataset pulled includes variables of importance such as product involved, product industry and affiliated codes, adverse symptoms reported, and overall outcomes based on severity. The temporal characteristics include report year with spacial characteristics summarized for the entire U.S. The product's role details whether it is generally suspect in the adverse events or if contamination was involved.See link to datasite here : https://open.fda.gov/apis/download-api-fields/***

## 3. The FDA's adverse event site reports adverse events for different industries. Summarize into a dataset adverse events for a **sample of 100 reports** by year from 2004-2023 for the Nut product industry. Perform some wrangling of needed data columns including important information such as adverse events, product name, industry name and code, product's role in adverse event, and observed reactions.

*Hint: The industry code is 23 and you will need to use the function "GET()" to pull the right API dataset.*

```{r FDA API link}
#| warning: false
events <- GET("https://api.fda.gov/food/event.json?search=date_started:[20040101+TO+20230101]") 


```

```{r API for adverse events from edible nuts}
#| warning: false
nuts <- GET("https://api.fda.gov/food/event.json?search=products.industry_code:23&limit=100")

summary(nuts)

```

```{r converting data into character vector from JSON format}
#| warning: false
new_nuts <- fromJSON(rawToChar(nuts$content))

#head(new_nuts)

```

```{r unnested nutella}
#| warning: false

nutella <- new_nuts[["results"]] %>%
  select(date_created, outcomes, reactions, products)

#head(nutella)
```

```{r unnesting nutella the right way}
nutella1 <- nutella %>%
  unnest_longer(outcomes, values_to = "outcomes") %>%
  unnest_longer(reactions, values_to = "reactions")
  

#head(nutella1)
```

```{r}
#| warning: false
nutella2 <- nutella1 %>%
  unnest_wider(products, names_sep = "_" )

head(nutella2)
```

## 4. How many adverse events for nut products were reported each year between the years of 2004-2023? Summarize findings in a bar chart showing total events per year.

```{r year summary}
#| warning: false
nutella_year <- nutella2 %>%
  separate(date_created, into = c("year", "month_day"), sep = 4)
#nutella_year

nutella_year2 <- nutella_year %>%
  separate(month_day, into = c("month", "day"), sep = 2)

glimpse(nutella_year2)


```

```{r plotting bar chart for total events}
#| warning: false

year_nutsum <- nutella_year2 %>%
  group_by(year) %>%
count(year, name = "total_events")

year_nutsum

ggplot(year_nutsum, aes(x = factor(year), y = total_events)) + 
  geom_col(fill = "forestgreen") + 
  labs(title = "Adverse Events for Nut Products in U.S. by Year",
       x = "Year",
       y = "Total Adverse Events"
       ) + 
  theme_bw()


```

## 5. What products were involved in each year's total adverse events? Summarize results in a bar chart for each year showing total adverse events for each product involved in that given year. Show just one plot for 2010 for total adverse events .

*Hint: use a mapping function to change the plot name by year for data used in it.*

```{r product year summary}
#| warning: false
product_events <- nutella_year2 %>%
  count(year, products_name_brand, name = "Events_by_Product")

product_events$products_name_brand[[1]]
length(product_events$products_name_brand[[1]])

#so trying to run below chunk for mapping didn't work with this column being the character type it is so I have to convert it below. It's a list so maybe I can apply a functoin that works over a list like lapply or sapply : use paste() for joining list elements to create one name for each row and then adding a "_" to join names separated originally in the list. 

product_events <- product_events %>%
  mutate(products_name_brand = sapply(products_name_brand, function(x) paste(x, collapse = "_" )))

product_events

```

```{r plot product events, use mapping to change year plot name}
#| warning: false
year_of_nutella <- sort(unique(product_events$year))


nut_plots <- map(year_of_nutella, ~ {nutty_year <- product_events %>% filter(year == .x)

ggplot(nutty_year, aes(x = products_name_brand, y = Events_by_Product, fill = products_name_brand)) + 
  geom_col() +
  labs(title = paste("Adverse Events By Nut Product ", .x),
       x = "Nut Product", 
       y = "Total Adverse Events") +
  theme_bw() +
  #adding next line because when I previewed plots you the x-axis labels were overlapping too much to read: 
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        legend.position = "none")

})

names(nut_plots) <- year_of_nutella


nut_plots$`2010`

#omg it actually worked
```

## 6. For each nut product associated with adverse events, summarize in a bar plot the total number of outcomes for each outcome type reported. What products resulted in hospitalization? 
*Hint: you will need to set up y-axis to have total number of outcome reports, with bars for each product on x-axis by outcome type.*

```{r creating data frame for outcomes and character type product name}
#| warning: false

product_outcome <- nutella_year2 %>%
  mutate(products_name_brand = sapply(products_name_brand, function(x) paste(x, collapse = "_" ))) %>%
  count(products_name_brand, outcomes, name = "Net_outcomes")

product_outcome

```

```{r making one nice bar chart for outcomes}
#| warning: false
#adding below mod because of issues rendering plots

#will use split() function here instead of filter() like previously to split dataframe based on grouping variable "outcome". Then I'll try to use mapping like I did before. 

outcome_plots <- split(product_outcome, product_outcome$outcomes)
outcome_plots

```

```{r mapping over split up data frames for plots}
# note for meghan: .x in function() below is a placeholder for each data frame that was split by outcome in previous chunk. 

nut_plot3 <- map(
  outcome_plots, 
  function(.x) {nut_outcome <- unique(.x$outcomes)
  ggplot(.x, aes(x = products_name_brand, 
                y = Net_outcomes)) +
    geom_col(fill = "purple") + 
    #had to add bottom part because titles kept being cut off in test plots 
   # coord_flip(clip = "off") + 
    #above didn't work so I just turned the flip off. 
    labs( title = paste("Products Associated with Certain Adverse Outcomes:", nut_outcome),
          x = "Product Name",
          y = "Total Outcome Events") + 
    theme_bw() +
    theme(
      axis.text.y = element_text(size = 5),
      axis.text.x = element_text(angle = 76, hjust = 1),
      plot.margin = margin(t = 80, r = 80, b = 20, l = 20))
  
  }
)

nut_plot3$Hospitalization

```
***I tried several times to fit the plots above into a correct window size but as you can see I may need help troubleshooting this. ***


## 7. Show using a plot the highest to lowest years for "Visited a healthcare provider" outcomes for nut products from the 2004-2023 dataset. 

```{r filter some outcomes}

provider_1 <- nutella_year2 %>%
  filter(outcomes == "Visited a Health Care Provider") %>%
 count(year, name = "Provider_visits") %>%
  arrange(desc(Provider_visits))

provider_1

```
```{r question 7 plotting after ranking}
#| warning: false
ggplot(provider_1, aes(x = reorder(as.factor(year), Provider_visits), y = Provider_visits)) + 
  geom_col(fill = "red") + 
  labs(
    title = "Healthcare Provider Visit Outcomes By Year",
    x = "Year",
    y = "Number of Events" 
    
  ) + 
  theme_dark()


```


## 8. Which product had the most adverse events of all time? How many adverse events were associated with this product? 

```{r max adverse events}
#| warning: false
#let me use the dataset i made earlier already listing the total number of adverse events per product
product_events

worst_product <- product_events %>%
filter(Events_by_Product == max(Events_by_Product, na.rm = TRUE)) %>%
  pull(products_name_brand, Events_by_Product)

worst_product
```
***The product with the most adverse events within the nut product industry was JIF Peanut Butter with 28 reported events between 2004-2023. ***


Sources used: https://stackoverflow.com/questions/64883971/r-return-column-name-for-row-where-first-given-value-is-found, Perplexity Best Model using the following sources: https://dplyr.tidyverse.org/reference/index.html, https://ggplot2.tidyverse.org/reference/, https://r-graph-gallery.com/267-reorder-a-variable-in-ggplot2, https://aosmith.rbind.io/2018/08/20/automating-exploratory-plots/, https://www.geeksforgeeks.org/r-language/apply-lapply-sapply-and-tapply-in-r/, https://rfortherestofus.com/2023/11/pivot-functions, 

**Personal Feedback: I found the question-making part of this exercise really  helpful for thinking about how I would set up the code and what functions I know I would want to use. The biggest struggle in completing this exercise was figuring out aspects of ggplot which wouldn't work if I changed my window size, etc. I also found navigating the open-source data difficult and had to research a lot of tutorials for wrangling the data in the format it came in. **


# Submitting your work

Once you have developed all the code and answers, make sure to Render this quarto file.

**Notes on rendering**:

-   Make sure to render your work and inspect how the final html look like.\
-   If it does not look professional for whatever reason, then fix the issue, re-render it, recheck.\
-   Only send me your work once your html file looks professional.
    -   **DO NOT** delete the file's heading levels (# and ##). They set up the proper heading 1 and 2 levels, and I use them to guide my grading.\

    -   If a given chunk is also outputting warnings or messages, inhibit this behavior by changing the chunk options `message` and `warning` to `FALSE` (I'll demo this in a moment).

    -   If, after rendered, 2 lines of text are connected and you wish to "break line" between them, add 2 extra spaces after the first one (I'lld demo this in a moment).

After rendering, an .html file will be created on your `code` folder.

Rename this file to `Assignment-06-opendata-LASTNAME.html`.

For ex., mine would be `Assignment-06-opendata-Bastos.html`.

Submit your work by uploading the **html** file to **eLC Assignment #6 - open data** by Monday March 16 11:59 pm.
